Climate-driven phytoplankton community variability in the Cariaco Basin, Venezuela

### 2025-05-13 agujournaltemplate.tex
added new ZScore plot to manuscript
editing tables & new figures

\begin{figure}
\noindent\includegraphics[width=\textwidth]{fig/PLOTZScores_updated.pdf}
\caption{Heatmap of yearly average values, normalized for comparison as z-scores. Red indicates a larger value, blue indicates a smaller value for that particular year. The years 1995 (n=2) and 2017 (n=1) were removed due to lack of data, additionally we marked years with two or less and three or more missing values. Diversity Indices are calculated based on microscopy counts at the Genus level.}
\label{fig:zscore}
\end{figure}

@@

\begin{figure}
\noindent\includegraphics[width=\textwidth]{fig/ClusteringCompPlot_NEW 2.pdf}
\caption{(a) Clustering of microscopy data at the Genus level, using the binary Jaccard distance. The community for each year is binned together, to smooth out seasonal aspects. (b) Species turnover between years for annual aggregate of identified phytoplankton to a genus level. The three genera dominating the cell counts for the separate parts of cluster 1 and cluster 2 are given in order of abundance from top to bottom. (c) NMDS plot of monthly community data (cell counts) that is cube-root transformed. Stress: 0.261, 2 dimensions, converging solution. Color of dots based on which yearly cluster the monthly data is in. Vector overlay of ENVFIT shows underlying environmental variables that are co-varying with the community data.
Plots (d-o) below show density distributions of the environmental variables as well as diversity indices between the clustered years. Dashed line shows the median and significance levels of Wilcoxon sum rank tests between clusters are shown above plots  (***\textless0.001, **\textless0.01, *\textless0.05, n.s.=not significant).}
\label{fig:clustering}
\end{figure}

@@

\caption{Individual Gradient Forest model runs for each variable and the corresponding time lags. 
        For in-situ variables, for which no coverage extends over the time series, 
        we tested all measurements up to a lag of 3 months. 
        For climate variables, we tested up to a lag of 6 months. 
        Time lag with maximum importance per variable are highlighted in bold and chosen for the final model run.}

\ref{app:table1}
\centering
\begin{tabular}[t]{rrrrrrrrrr}
\toprule
lag & NO3 & PO4 & SiO4 & Temperature & Isotherm\_21 & Salinity\_bottles & u10 & AMO & MEIv2\\
\midrule
0 & \textbf{0.038} & \textbf{0.038} & \textbf{0.034} & \textbf{0.071} & \textbf{0.028} & 0.007 & 0.004 & 0.016 & 0.015\\
1 & 0.017 & 0.011 & 0.015 & 0.002 & 0.002 & 0.004 & \textbf{0.010} & 0.014 & 0.018\\
2 & 0.022 & 0.008 & 0.013 & 0.006 & 0.008 & \textbf{0.011} & 0.009 & \textbf{0.030} & 0.019\\
3 & 0.016 & 0.007 & 0.004 & 0.005 & 0.007 & 0.010 & 0.007 & 0.018 & 0.018\\
4 & - & - & - & - & - & - & 0.008 & 0.019 & \textbf{0.029}\\
5 & - & - & - & - & - & - & 0.008 & 0.018 & 0.022\\
6 & - & - & - & - & - & - & 0.010 & 0.018 & 0.013\\
\bottomrule
\end{tabular}
\end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Optional Glossary, Notation or Acronym section goes here:
%
% Glossary is only allowed in Reviews of Geophysics
%  \begin{glossary}
%  \term{Term}
%   Term Definition here
%  \term{Term}
%   Term Definition here
%  \term{Term}
%   Term Definition here
%  \end{glossary}

>>>>

\begin{figure}
\noindent\includegraphics[width=\textwidth]{fig/Figure2_ZScores_v2.pdf}
\caption{Heatmap of yearly average values, normalized for comparison as z-scores. Red indicates a larger value, blue indicates a smaller value for that particular year. The years 1995 (n=2) and 2017 (n=1) were removed due to lack of data, additionally we marked years with two or less and three or more missing values. Diversity Indices are calculated based on microscopy counts at the Genus level.}
\label{fig:zscore}
\end{figure}

@@

\begin{figure}
\noindent\includegraphics[width=\textwidth]{fig/Figure4_ClusteringNMDS_v2.pdf}
\caption{(a) Clustering of microscopy data at the Genus level, using the binary Jaccard distance. The community for each year is binned together, to smooth out seasonal aspects. (b) Species turnover between years for annual aggregate of identified phytoplankton to a genus level. The three genera dominating the cell counts for the separate parts of cluster 1 and cluster 2 are given in order of abundance from top to bottom. (c) NMDS plot of monthly community data (cell counts) that is cube-root transformed. Stress: 0.261, 2 dimensions, converging solution. Color of dots based on which yearly cluster the monthly data is in. Vector overlay of ENVFIT shows underlying environmental variables that are co-varying with the community data.}
\label{fig:clustering}
\end{figure}

\begin{figure}
\noindent\includegraphics[width=\textwidth]{fig/Figure5_ClustCompPlot_v2.pdf}
\caption{(Plots (a-o) below show density distributions of the environmental variables as well as diversity indices between the clustered years. Dashed line shows the median and significance levels of Wilcoxon sum rank tests between clusters are shown above plots  (***\textless0.001, **\textless0.01, *\textless0.05, n.s.=not significant).}
\label{fig:clustcomp}
\end{figure}

@@

\caption{Individual Gradient Forest model runs for each variable and the corresponding time lags. 
        For in-situ variables, for which no coverage extends over the time series, 
        we tested all measurements up to a lag of 3 months. 
        For climate variables we tested up to a lag of 6 months. 
        Time lag with maximum importance per variable are highlighted in bold and chosen for the final model run.}
\centering
\resizebox{\textwidth}{!}{\begin{tabular}[t]{rrrrrrrrrrrr}
\toprule
lag & NO_3 & PO_4 & SiO_4 & SST & \qty{21}{\celsius} isotherm & Salinity & Wind speed & AMO & MEI v.2 & Precipitation & Evaporation\\
\midrule
0 & \textbf{0.025} & \textbf{0.028} & \textbf{0.029} & \textbf{0.084} & \textbf{0.024} & 0.006 & 0.003 & 0.015 & 0.015 & 0.004 & 0.010\\
1 & 0.013 & 0.011 & 0.014 & 0.010 & 0.003 & 0.007 & \textbf{0.007} & 0.012 & 0.015 & \textbf{0.013} & 0.011\\
2 & 0.016 & 0.007 & 0.013 & 0.011 & 0.006 & 0.011 & 0.007 & \textbf{0.026} & 0.018 & 0.005 & 0.009\\
3 & 0.014 & 0.007 & 0.005 & 0.007 & 0.007 & \textbf{0.016} & 0.004 & 0.017 & 0.017 & 0.008 & \textbf{0.013}\\
4 & NA & NA & NA & NA & NA & NA & 0.006 & 0.018 & \textbf{0.025} & 0.008 & 0.005\\
\addlinespace
5 & NA & NA & NA & NA & NA & NA & 0.007 & 0.016 & 0.020 & 0.008 & 0.032\\
6 & NA & NA & NA & NA & NA & NA & 0.006 & 0.017 & 0.011 & 0.006 & 0.003\\
12 & NA & NA & NA & NA & NA & NA & NA & 0.011 & 0.012 & NA & NA\\
\bottomrule
\end{tabular}}
\end{table}

\caption{Individual Gradient Forest model runs for each variable and the corresponding time lags. 
        For in-situ variables, for which no coverage extends over the time series, 
        we tested all measurements up to a lag of 3 months. 
        For climate variables we tested up to a lag of 6 months. 
        Time lag with maximum importance per variable are highlighted in bold and chosen for the final model run.}
\centering
\resizebox{\textwidth}{!}{\begin{tabular}[t]{rrrrrrrrrrrr}
\toprule
lag & NO3 & PO4 & SiO4 & Temperature & Isotherm\_21 & Salinity\_bottles & u10 & AMO & MEIv2 & Precipitation & Evaporation\\
\midrule
0 & 0.062 & 0.11 & 0.068 & 0.092 & 0.067 & 0.175 & 0.048 & 0.013 & 0.082 & 0.048 & 0.019\\
1 & 0.148 & NA & 0.017 & 0.019 & 0.044 & NA & 0.158 & 0.033 & 0.057 & 0.094 & 0.346\\
2 & 0.177 & NA & NA & NA & 0.013 & NA & NA & 0.057 & 0.140 & 0.254 & 0.116\\
3 & 0.341 & NA & NA & NA & NA & NA & NA & 0.023 & 0.143 & NA & 0.016\\
4 & NA & NA & NA & NA & NA & NA & NA & 0.081 & 0.048 & NA & 0.209\\
\addlinespace
5 & NA & NA & NA & NA & NA & NA & NA & 0.260 & 0.095 & NA & NA\\
6 & NA & NA & NA & NA & NA & NA & NA & NA & 0.046 & NA & NA\\
12 & NA & NA & NA & NA & NA & NA & NA & NA & 0.037 & NA & NA\\
\bottomrule
\end{tabular}}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Optional Glossary, Notation or Acronym section goes here:
%
% Glossary is only allowed in Reviews of Geophysics
%  \begin{glossary}
%  \term{Term}
%   Term Definition here
%  \term{Term}
%   Term Definition here
%  \term{Term}
%   Term Definition here
%  \end{glossary}
